---
title: "Homologous gene enrichment"
author: "Antoine Beauchamp"
date: '2022-07-25'
output: 
  html_document:
    theme: paper
    highlight: pygments
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tmod))
```

# Import gene sets

```{r}
# Path to gene set data
annotation_file_remote <- 
  paste0("http://download.baderlab.org/EM_Genesets/",
         "November_01_2019/Mouse/symbol/GO/MOUSE_GO_bp_no_GO_iea_symbol.gmt")

#Output file
annotation_file_local <- "../data/MOUSE_GO_bp_no_GO_iea_symbol.gmt"

# Download it if it doesn't exist
if (!file.exists(annotation_file_local)) {
  download.file(annotation_file_remote, annotation_file_local)
}

# Load GMT file into a tmod object
mappings <- tmodImportMSigDB(file = annotation_file_local, format = "gmt")

homologous_genes <- read_csv('../data/MouseHumanGeneHomologs.csv', 
                             show_col_types = FALSE) %>% 
  pull(Mouse)
```

# Enrichment in one module


Here we consider the module "Nervouss System Development". This module has an associated gene set. 

We can examine what number of homologous genes fall into the Nervous System Development module. This module has 2237 genes. 

```{r}
#Number of genes in the module
B <- mappings$MODULES2GENES[["NERVOUS SYSTEM DEVELOPMENT%GOBP%GO:0007399"]] %>% 
  length 
print(B)
```
There are 2237 genes associated with this module. We want to identify the number of genes in this module that are also part of the homologous gene subset: 

```{r}
#Number of homologous genes in the module
b <- homologous_genes %>% 
  intersect(mappings$MODULES2GENES[["NERVOUS SYSTEM DEVELOPMENT%GOBP%GO:0007399"]]) %>% 
  length 
print(b)
```
There are 709 homologous genes in this module.

```{r}
b/B
```
This corresponds to a proportion of about 32%. Now how many genes are in the full annotated set? And how many of those genes are also in the homologous gene set? 

```{r}
# Number of genes in the entire set
N <- mappings$GENES %>% 
  pull(ID) %>% 
  length 
print(N)
```

```{r}
# Number of homologous genes in the entire set
n <- homologous_genes %>% 
  intersect(mappings$GENES$ID) %>% 
  length 
print(n)
```
```{r}
n/N
```

This proportion is about 17%, rather than 32%. We can examine over-representation of the homologous genes in the module set by taking the quotient of these ratios. The idea is to compare the proportion of homologous genes in the module set to the proportion of homologous genes in the entire set. If there is no relationship between the homologous gene set and the module set, we expect the proportions to be the same and the ratio to be 1. 

```{r}
E <- (b/B)/(n/N) # ratio of proportions
print(E)
```
We find that the homologous genes are over-represented in the module set that we specified. However there is some statistical variation to this, so we need to compare it to a null distribution. The null distribution for a ratio of this kind is the hypergeometric distribution. Sampling the distribution 10,000 times, we can examine where our true value falls compared with the null. 

```{r}
tibble(x = rhyper(10000, B, N-B, n)) %>% 
  ggplot(aes(x = x)) + 
  geom_histogram(col = 'black',
                 fill = 'grey70',
                 binwidth = 5) +
  geom_vline(xintercept = b,
             col = 'red',
             linetype = 'dashed') + 
  coord_cartesian(xlim = c(300, 750)) + 
  theme_bw()
```
Our true value is quite clearly far outside the tails of the distribution. The associated p-value is: 

```{r}
phyper(b-1, B, N-B, n, lower.tail = F)
```
So we can say that the homologous genes are indeed over-represented in the given module. 

# Enrichment in all modules

The analysis above was only for the module termed "nervous system development". We would like to repeat this process for every module in the data set. This can be done automatically using `tmodHGtest()` 

This was only for the nervous system development module. We want to repeat this process for every module in our set. 

```{r}
#Run the test for every module
results <- tmodHGtest(fg = homologous_genes,
                      bg = mappings$GENES$ID,
                      mset = mappings,
                      qval = 0.05,
                      order.by = 'pval',
                      filter = T)

results_tbl <- results %>% 
  as_tibble() %>% 
  separate(ID, sep = '%', into = c('Name', 'Source', 'Identifier')) %>% 
  select(Identifier, Title, b, B, n, N, E, P.Value, adj.P.Val) %>% 
  arrange(adj.P.Val)

results_tbl
```


```{r}
results_tbl %>% 
  filter(adj.P.Val <= 0.001) %>% 
  nrow()
```



