---
title: "Exploring the multi-layer perceptron using integrated gradients"
author: "Antoine Beauchamp"
date: '2022-07-19'
output: 
  html_document:
    theme: paper
    highlight: pygments
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(data.tree))
suppressPackageStartupMessages(library(RMINC))
suppressPackageStartupMessages(library(MRIcrotome))
suppressPackageStartupMessages(library(patchwork))
```

# Functions

```{r}
source('../functions/tree_tools.R')

#' Plot top genes
#'
#' @param x (data.frame) Data frame containing summary statistics
#' @param n (integer scalar) Number of genes to plot
#'
#' @return ggplot
plot_top_genes <- function(x, n) {
  
  x %>% 
    top_n(n = n, wt = x_mean) %>% 
    ggplot(aes(x = Gene)) + 
    geom_pointrange(aes(y = x_mean,
                        ymin = x_sd_lower,
                        ymax = x_sd_upper)) + 
    coord_flip() + 
    labs(y = 'Feature importance',
         title = str_c('Top genes for region: ', struct),
         caption = 'Mean +/- 95% CI over latent spaces') + 
    theme_bw() + 
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
}

#' Subset top genes
#'
#' @param x (data.frame) Data frame containing summary statistics
#' @param n (integer scalar) Number of genes to subset
#' @param homologs (data.frame) Data frame containing homologs
#'
#' @return (data.frame) Data frame containing mouse and human gene names
#' for top genes
select_top_genes <- function(x, n, homologs) {
  x %>% 
    top_n(n = n, wt = x_mean) %>% 
    arrange(desc(x_mean)) %>% 
    left_join(homologs, 
              by = c('Gene' = 'Human')) %>% 
    select(Mouse, Human = Gene) %>% 
    return()
}

#' Import gene ISH images
#'
#' @param genes (character vector) Mouse gene names
#' @param exprdir (character scalar) Paths to directory containing ISH images
#' @param mask (numeric vector) Image mask)
#'
#' @return (list) List containing mincSingleDim expression arrays
import_gene_images <- function(genes, exprdir, mask) {
  expr_files <- list.files(exprdir, full.names = TRUE)
  gene_arrays <- vector(mode = 'list', length = length(genes)) 
  for (i in 1:length(genes)) {
    infile <- str_subset(expr_files, str_c(genes[i], '_'))
    img <- mincGetVolume(infile)
    img[mask == 0] <- 0
    img[img == -1] <- 0
    gene_arrays[[i]] <- img
  }
  names(gene_arrays) <- genes
  return(gene_arrays)
}

#' Convert data frame columns to images
#'
#' @param genes (character vector) Names of columns to use
#' @param expr (data.frame) Data frame containing voxel-wise values
#' @param mask (mincSingleDim) Mask identifying which voxels correspond to data frame rows
#'
#' @return (list) List containing mincSingleDim expression arrays
columns_to_images <- function(genes, expr, mask) {
  gene_arrays <- vector(mode = 'list', length = length(genes))
  for (i in 1:length(genes)) {
    img <- numeric(length(mask))
    img[mask == 1] <- pull(expr[,genes[i]])
    attributes(img) <- attributes(mask)
    gene_arrays[[i]] <- img
  }
  names(gene_arrays) <- genes
  return(gene_arrays)
}

#' Create a slice series for multiple images
#'
#' @param images (list) List of mincSingleDim arrays to visualize
#' @param anat (mincSingleDim) Anatomy template
#' @param nslices (integer scalar) Number of slices to plot per image
#' @param begin (integer scalar) Starting slice
#' @param end (integer scalar) Ending slice
#' @param low (numeric scalar) Lower threshold for overlay
#' @param high (numeric scalar) Upper threshold for overlay
#' @param symmetric (logical scalar) Whether to visualize symmetric
#' @param legend_title (character scalar)
#'
#' @return (environment) Slice series
top_gene_slice_series <- function(images, anat, nslices = 8, begin = 30, end = 50, low, high, symmetric = FALSE, legend_title = NULL) {
  
  ss_base <- sliceSeries(nrow = nslices, ncol = 1, begin = begin, end = end) %>% 
    anatomy(mincArray(anat), low = 700, high = 1400) 
  
  for (i in 1:length(images)) {
    if (i == 1) {
      ss <- ss_base
    } else {
      ss <- ss %>% 
        sliceSeries() %>% 
        anatomy() 
    }
    ss <- ss %>% 
      overlay(mincArray(images[[i]]), low = low, high = high, symmetric = symmetric) %>% 
      addtitle(names(images)[i])
  }
  ss <- ss %>% 
    legend(legend_title)
  
  return(ss)
}
```

# Importing general purpose data

```{r message=FALSE}
#Mouse-human homologous genes
homologs <- read_csv('../data/MouseHumanGeneHomologs.csv', 
                     show_col_types = FALSE)

#Path to AMBA ISH directory
exprdir <- '../AMBA/data/expression/coronal/'

#Processed expression matrix
expr_file <- '../data/MouseExpressionMatrix_voxel_coronal_maskcoronal_log2_grouped_imputed_labelled_scaled.csv'
df_expr <- data.table::fread(expr_file, header = TRUE) %>% 
  as_tibble()

#Mouse tree
load('../AMBA/data/MouseExpressionTree_DSURQE.RData')
tree_mouse <- Clone(treeMouseExpr)
rm(treeMouseExpr)

pruneAnatTree(tree_mouse,
              nodes = c("fiber tracts", "ventricular systems"),
              method = "AtNode")

#Imaging files
anatfile <- '../AMBA/data/imaging/DSURQE_CCFv3_average_200um.mnc'
maskfile <- '../AMBA/data/imaging/coronal_200um_coverage_bin0.8.mnc'
labelfile <- '../AMBA/data/imaging/DSURQE_CCFv3_labels_200um.mnc'

anat <- mincGetVolume(anatfile)
mask <- mincGetVolume(maskfile)
labels <- mincGetVolume(labelfile)

#Filters to map matrix back to imaging space
ind_mask <- mask == 1
ind_gm <- labels %in% tree_mouse[['Basic cell groups and regions']][['label']]
ind_filter <- ind_mask & ind_gm

mask_gm <- numeric(length(mask))
mask_gm[ind_filter] <- 1
attributes(mask_gm) <- attributes(mask)

#Create a 67-label atlas
load('../data/TreeLabelsReordered.RData')
tree_mouse_67 <- Clone(tree_mouse)
pruneAnatTree(tree_mouse_67,
              nodes = listLabelsMouseReordered$Region67_reordered,
              method = 'BelowNode')
labels_67 <- hanatToAtlas(anatTree = tree_mouse_67, labelVolume = mincArray(labels))
defs_67 <- hanatToAtlasDefs(anatTree = tree_mouse_67)
```


# Importing integrated gradients data

```{r}
#Identify files containing integrated gradients
files <- list.files('../data/MLP_outcomes/', full.names = T) %>% 
  str_subset('IntegratedGradients')

#Extract latent space IDs
latent_space_ids <- files %>% 
  str_extract('IntegratedGradients_[0-9]+.csv') %>% 
  str_extract('[0-9]+') %>% 
  as.integer()

#Name files according to latent space IDs
names(files) <- latent_space_ids

#How many files at time of rendering?
print(length(files))
```

```{r}
#Import gradients data
df_integrated_grads <- map_dfr(.x = files,
                               .f = read_csv,
                               .id = 'LatentSpace',
                               show_col_types = FALSE) %>% 
  mutate(LatentSpace = as.integer(LatentSpace)) %>% 
  arrange(LatentSpace)
```

```{r message = FALSE}
#Convert data to long format
df_integrated_grads_long <- df_integrated_grads %>% 
  pivot_longer(cols = c(-Region, -LatentSpace), names_to = 'Gene', values_to = 'x')

#Compute gradient summaries over latent spaces
df_integrated_grads_summary <- df_integrated_grads_long %>% 
  group_by(Region, Gene) %>% 
  summarise(x_mean = mean(x),
            x_sd = sd(x),
            x_sd_lower = x_mean - 2*x_sd,
            x_sd_upper = x_mean + 2*x_sd) %>% 
  ungroup()

#Split data frames according to regions
integrated_grads_structs <- split(x = df_integrated_grads_summary,
                                  f = factor(df_integrated_grads_summary$Region))

#Order genes according to importance
integrated_grads_structs <- integrated_grads_structs %>% 
  map(function(x) {mutate(x, Gene = factor(Gene, levels = Gene[order(x_mean)]))})
```

# Overview of gene importances

```{r fig.height = 8, fig.width = 10}
for (i in 1:length(integrated_grads_structs)) {
  plt <- ggplot(integrated_grads_structs[[i]],
                aes(x = as.numeric(Gene), y = x_mean)) + 
    geom_area(stat = 'identity', fill = 'grey70') + 
    geom_hline(yintercept = 0) + 
    geom_point(size = 0.5) + 
    labs(x = 'Genes',
         y = 'Feature importance',
         title = names(integrated_grads_structs)[i]) + 
    theme_bw() + 
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
  
  if (i == 1) {
    overview_plot <- plt
  } else {
    overview_plot <- overview_plot / plt
  }
}

overview_plot + 
  plot_annotation(title = 'Integrated gradients averaged over latent spaces')
```

# Digging into the caudoputamen


```{r}
struct <- 'Caudoputamen'
df_struct <- integrated_grads_structs[[struct]]

struct_label <- defs_67 %>% 
  filter(Structure == struct) %>% 
  pull(Label)

mask_struct <- numeric(length(labels_67))
mask_struct[labels_67 == struct_label] <- 1
attributes(mask_struct) <- attributes(labels_67)
```

```{r fig.width = 10, fig.height = 5}
plot_top_genes(x = df_struct, n = 10)
```

```{r}
ngenes <- 5
top_genes <- select_top_genes(x = df_struct,
                              n = ngenes,
                              homologs = homologs)

top_genes_human <- pull(top_genes, Human)
top_genes_mouse <- pull(top_genes, Mouse)
```

First we'll take a look at the in-situ hybridization images, normalized using the sum of the expression over the brain. 

```{r fig.width = 10, fig.height = 10}
#Import gene images
gene_maps <- import_gene_images(genes = top_genes_mouse, 
                                exprdir = exprdir, 
                                mask = mask)

#Normalize using total intensity
gene_maps_normalized <- map(.x = gene_maps,
                            .f = function(x){x/sum(x)})

ss_gene_maps_normalized <- top_gene_slice_series(images = gene_maps_normalized, anat = anat,
                                                 nslices = 8, begin = 30,end = 50,
                                                 low = 0.0001, high = 0.001,
                                                 symmetric = F, legend_title = 'Expression')

ss_gene_maps_normalized %>% 
  sliceSeries(nrow = 8, ncol = 1, begin = 30, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```

Not seeing anything for Prrg2 at this threshold. However this isn't what the perceptron is seeing as input. Let's instead taking a look at the processed and normalized expression values that are passed to the network. 

```{r fig.width = 10, fig.height = 10}
gene_maps_normalized <- columns_to_images(genes = top_genes_human,
                                          expr = df_expr,
                                          mask = mask_gm)
names(gene_maps_normalized) <- top_genes_mouse

ss_gene_maps_normalized <- top_gene_slice_series(images = gene_maps_normalized, anat = anat,
                                                 nslices = 8, begin = 30,end = 50,
                                                 low = 1, high = 3,
                                                 symmetric = T, legend_title = 'Normalized expression')

ss_gene_maps_normalized %>% 
sliceSeries(nrow = 8, ncol = 1, begin = 30, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```

I suppose that makes more sense?

Prrg2 has a stronger signal here. But I wonder how much of that is true signal and how much is imputed. 

```{r fig.width = 6, fig.height = 8}
Prrg2 <- gene_maps[['Prrg2']]
Prrg2_log <- Prrg2
Prrg2_log[Prrg2_log <= 0] <- NA
Prrg2_log <- log2(Prrg2_log)
Prrg2_log[is.na(Prrg2_log)] <- 0

sliceSeries(nrow = 8, ncol = 1, begin = 30, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(Prrg2), low = 0.001, high = 0.5, symmetric = F) %>% 
  legend('Expression') %>% 
  sliceSeries() %>% anatomy() %>% 
  overlay(mincArray(Prrg2_log), low = 2, high = 10, symmetric = T) %>% 
  legend('log2(Expression)') %>% 
  sliceSeries() %>% anatomy() %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```

Looks like the signal is real. 


# Digging into the primary motor area


```{r}
struct <- 'Primary motor area'
df_struct <- integrated_grads_structs[[struct]]

struct_label <- defs_67 %>% 
  filter(Structure == struct) %>% 
  pull(Label)

mask_struct <- numeric(length(labels_67))
mask_struct[labels_67 == struct_label] <- 1
attributes(mask_struct) <- attributes(labels_67)
```

```{r fig.width = 10, fig.height = 5}
plot_top_genes(x = df_struct, n = 10)
```

```{r}
ngenes <- 5
top_genes <- select_top_genes(x = df_struct,
                              n = ngenes,
                              homologs = homologs)

top_genes_human <- pull(top_genes, Human)
top_genes_mouse <- pull(top_genes, Mouse)
```

```{r fig.width = 10, fig.height = 10}
#Import gene images
gene_maps <- import_gene_images(genes = top_genes_mouse, 
                                exprdir = exprdir, 
                                mask = mask)

#Normalize using total intensity
gene_maps <- map(.x = gene_maps,
                 .f = function(x){x/sum(x)})

ss_gene_maps <- top_gene_slice_series(images = gene_maps, anat = anat,
                                      nslices = 8, begin = 30,end = 50,
                                      low = 0.00001, high = 0.0001,
                                      symmetric = F, legend_title = 'Expression')

ss_gene_maps %>% 
sliceSeries(nrow = 8, ncol = 1, begin = 30, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```


Maybe need to take into account normalization?


```{r fig.width = 10, fig.height = 10}
gene_maps_normalized <- columns_to_images(genes = top_genes_human,
                                          expr = df_expr,
                                          mask = mask_gm)
names(gene_maps_normalized) <- top_genes_mouse

ss_gene_maps_normalized <- top_gene_slice_series(images = gene_maps_normalized, anat = anat,
                                                 nslices = 8, begin = 30,end = 50,
                                                 low = 1, high = 3,
                                                 symmetric = T, legend_title = 'Normalized expression')

ss_gene_maps_normalized %>% 
sliceSeries(nrow = 8, ncol = 1, begin = 30, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```







# Digging into the infralimbic area


```{r}
struct <- 'Infralimbic area'
df_struct <- integrated_grads_structs[[struct]]

struct_label <- defs_67 %>% 
  filter(Structure == struct) %>% 
  pull(Label)

mask_struct <- numeric(length(labels_67))
mask_struct[labels_67 == struct_label] <- 1
attributes(mask_struct) <- attributes(labels_67)
```

```{r fig.width = 10, fig.height = 5}
plot_top_genes(x = df_struct, n = 10)
```

```{r}
ngenes <- 5
top_genes <- select_top_genes(x = df_struct,
                              n = ngenes,
                              homologs = homologs)

top_genes_human <- pull(top_genes, Human)
top_genes_mouse <- pull(top_genes, Mouse)
```

```{r fig.width = 10, fig.height = 10}
#Import gene images
gene_maps <- import_gene_images(genes = top_genes_mouse, 
                                exprdir = exprdir, 
                                mask = mask)

#Normalize using total intensity
gene_maps <- map(.x = gene_maps,
                 .f = function(x){x/sum(x)})

ss_gene_maps <- top_gene_slice_series(images = gene_maps, anat = anat,
                                      nslices = 8, begin = 42,end = 50,
                                      low = 0.00005, high = 0.0001,
                                      symmetric = F, legend_title = 'Expression')

ss_gene_maps %>%
sliceSeries(nrow = 8, ncol = 1, begin = 42, end = 50) %>%
  anatomy(mincArray(anat), low = 700, high = 1400) %>%
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>%
  addtitle('Target ROI') %>%
  draw()
```


Maybe need to take into account normalization?


```{r fig.width = 10, fig.height = 10}
gene_maps_normalized <- columns_to_images(genes = top_genes_human,
                                          expr = df_expr,
                                          mask = mask_gm)
names(gene_maps_normalized) <- top_genes_mouse

ss_gene_maps_normalized <- top_gene_slice_series(images = gene_maps_normalized, anat = anat,
                                                 nslices = 8, begin = 42,end = 50,
                                                 low = 1, high = 2,
                                                 symmetric = T, legend_title = 'Normalized expression')

ss_gene_maps_normalized %>% 
sliceSeries(nrow = 8, ncol = 1, begin = 42, end = 50) %>% 
  anatomy(mincArray(anat), low = 700, high = 1400) %>% 
  overlay(mincArray(mask_struct), low = 0, high = 1, col = 'magenta') %>% 
  addtitle('Target ROI') %>% 
  draw()
```





